Constraints on magic state protocols from the statistical mechanics of \\Wigner negativity

### 2021-07-16 cover-letter/cover-letter.tex
Cover letter plus summary updated. Minor edits in main text.

\nick{this is supposed to be a ``100-word concise and compelling argument''. Currently: 203 words}

@@

Our work establishes a connection between two fields within quantum information: quantum thermodynamics and quasi-probability representations of quantum physics.
This allows for important advances in the subfield of magic theory, an increasingly important area of quantum computation.

Primarily, we develop new techniques that provide analytical bounds on general magic state distillation processes. We show that our bounds are tighter than previously defined bounds, and importantly can be expressed in terms of the physical symmetries of the circuit in question.
We therefore provide novel and powerful constraints on the design of magic protocols.

The implications of our analysis are also important foundationally.
We develop a novel theory of statistical mechanics on quasi-distributions and we further construct entropic functions which attain negative values, a property which acts as a fundamental witness of non-classicality.

For the above reasons, we believe that our manuscript is of relevance to a wide audience of quantum physicists, and so is suitable for publication in Phys. Rev. X.

@@

\nick{this is supposed to be 250 words long. Currently: 243 words}

Quantum computers promise astronomical speed-ups for important computational tasks.
A practically necessary requirement to achieve such speed-ups for useful computational tasks is establishing a framework that ensures fault-tolerance, which is the ability to correct computational errors in the circuit successfully and reliably.
It is well-known that restricting to this framework, it is impossible to achieve quantum speed-ups, therefore we need to initiate the circuit with some low-noise input state that contains enough resources which can be used by the circuit to perform the desired computational task.

Creating this low-noise, highly resourceful state, called a ``magic state'', is a costly process, requiring in principle a quantum process that manipulates many highly noisy copies of the magic state in order to distill one with very low noise.
Our work addresses the question regarding the highest possible rate of low-noise magic states one could extract, given some copies of highly noisy magic states.
We provide fundamental bounds on this rate, based on the thermodynamic properties of the quantum process, which constraint the efficiency of any general quantum processes attempting to distill magic.
As our bounds depend on the thermodynamic properties of the process, they notably provide a way of assessing how physical properties affect the distillation process.

On our way to addressing this question, we develop a novel physical framework of thermodynamics, where \nick{microstates of a system can be quasi-probabilistic} and where entropies can attain negative values, which are fully meaningful as indicators of quantum computational speed-ups.

>>>>

@@

A fundamental problem is to understand how quantum physics admits higher degrees of order than possible in classical physics. One signature of this non-classical order is the appearance of `negative probabilities' in the Wigner representation of quantum systems. However statistical mechanics cannot be directly applied to quantify this order since the Boltzmann entropy is not well-defined when Wigner negativity arises. In this work we do two main things: we show how a well-defined statistical mechanics of Wigner negativity can be constructed, and then apply this framework to quantify the order in `magic states', which are the key sources of non-classicality in leading quantum computing models.

It is a major goal to properly understand the processing and `distillation' of magic states for quantum technologies. Upper bounds on magic state distillation protocols do exist and are based on abstract information-theoretic measures. Our methods outperform existing upper bounds in the literature, and to our knowledge, provide the first general framework to incorporate dependence on hardware physics, such as system Hamiltonians or temperatures. As such this approach is relevant to the challenge of designing and building quantum computers.

Our results are also of interest beyond quantum technologies. We show that recent research on the thermodynamics of small quantum systems can be extended to Wigner negativity/quasi-distributions. To our knowledge, we provide the first application of majorization (a fundamental concept in classical statistics) to quasi-probability distributions, and so open up the possibility of new tools to analyse non-classicality more generally. We show that sub-set of classical (R\'{e}nyi) entropies remain meaningful on quasi-distributions and the appearance of negative entropies is a witness of non-classicality in quantum systems. Finally, our analysis also raises novel questions relevant to work on classical statistical mechanics.

For the above reasons, we believe that our manuscript is of relevance to a wide audience of quantum physicists, and so is suitable for publication in PRX Quantum.

@@

\ddd{You can use the first paragraph, which is 100 words, for the 100 word summary. We can include the cover-letter too, which is not 100 words. Find a place to add it in and make it clear e.g. in the title, it's a cover letter. This way the editor should open it and read it.}

\nick{this is supposed to be ``three paragraphs, first should be self-contained summary within summary, second and third more extensive description'', should be at most 250 words long. Currently: 250 words}

A fundamental problem is to understand how quantum physics admits higher degrees of order than classical physics. Here we show how quantum-mechanical order can be described in terms of a rigorous statistical mechanics framework in which negative entropies arise. We then apply this framework to quantify the order in `magic states' - crucial resources that power quantum computers. Our methods outperform existing magic bounds, and provide the first general framework to incorporate dependence on hardware physics, such as system Hamiltonians or temperatures. This work is relevant both to quantum technologies and the fundamental challenge of quantifying non-classicality in quantum systems.

A signature of non-classical order is the appearance of `negative probabilities' in representations of quantum systems. However traditional statistical mechanics cannot be directly applied to negative probabilities. We first show how to extend statistical mechanics by using a more fundamental concept of order (`majorization') from statistical theory and then apply this to quantifying the order in magic states.

Quantum computers `burn' magic states as a form of fuel to provide exponential speed-ups for important tasks, such as in computational Chemistry. Our results show that this thermodynamic intuition can be made precise and that magic states, in a rigorous sense, carry `negative entropy' that is processed thermodynamically. Our methods outperform existing magic bounds in the literature, and to our knowledge, provide the first general framework to incorporate dependence on hardware physics, such as system Hamiltonians or temperatures. As such this approach is relevant to the challenge of designing and building quantum computers.

### 2021-07-16 ms.tex
Cover letter plus summary updated. Minor edits in main text.

Work towards fault-tolerance is increasingly bridging the gap between abstract theory and experiment. Extensive work is being done on error mitigation~\cite{jones_2012, Li_2017, Temme_2017, Endo_2018, McClean_2017}, and the incorporation of hardware physics into the theoretical models~\cite{Kandala_2019, holmes_resource_2019, Colless_2018, song2018quantum, Bravyi_2021}. For example, the XZZX code~\cite{bonilla_ataides_xzzx_2021} is a variant of the surface code that incorporates noise bias explicitly and has been shown to attain the hashing bound of random codes~\cite{Bennett_1996}.

In this work, we develop a framework to analyse magic state distillation protocols where explicit physics of the system is incorporated in the distillation bounds. This is achieved by considering how a given magic state protocol transforms a \emph{pair} of quantum states -- one being a noisy magic state, and the other a stabilizer state that is distinguished by the physics of the system (for example a state at a characteristic temperature, or the maximally mixed state) and acts as a reference state for the protocol. The magic distillation bounds can then be expressed in terms of the physics of the reference state (e.g.~free energy changes in the case of temperature). Such bounds are of potential interest in assessing how physical features such as temperature, noise biases or fixed-point structure associated with restricted gate-sets constrain distillation protocols~\cite{Tuckett_2019, Aliferis_2008, Stephens_2013, Li_2015, Babbush_2018, Guillaud_2019, Fowler_2019}.

@@

The phase-point operators form an orthogonal operator basis with respect to the Hilbert-Schmidt inner product, as discussed in Appendix~\ref{app:wigner}.
Therefore, any quantum state $\rho$ can be expressed as a linear combination of them, $\rho = \sum_{\bmz \in \cal{P}_d} \W[\bmz]{\rho} A_{\bmz}$, where the coefficients $\W{\rho}(\z)$ give the Wigner distribution of state $\rho$,
\begin{equation}\label{eq:wstate}
    \W[\bmz]{\rho} \coloneqq \frac{1}{d}\tr[A_{\bmz} \rho].
\end{equation}

For any quantum state $\rho$, the Wigner distribution $W_\rho(\bmz)$ is readily seen to be a $d^2$-dimensional quasi-distribution over $\cal{P}_d$. More precisely, $W_\rho(\bmz)$ is a real-valued function on $\P_d$ with the normalisation property $\sum_{\bmz} W_\rho(\bmz) = 1$ (see Appendix~\ref{app:wigner} for details).

@@

\nick{We note that the discrete Wigner representation described in this section is in fact the only non-contextual quasi-probability representation of odd-dimensional stabilizer theories~\cite{schmid2021noncontextual}. In particular, negativity within the Wigner representation cannot be eliminated by considering a different quasi-probability representation. - look at paragraph starting ``In the setting of the full stabilizer subtheory...'' on page 2 of~\cite{schmid2021noncontextual}}

@@

Within the Wigner representation for odd prime dimension, it is well-known that all positively represented states used in Clifford circuits admit an efficient classical description~\cite{cit:mari}, so negativity is a necessary resource for universal fault-tolerant quantum computing~\cite{cit:veitch}. Therefore, the free states in any magic theory are required to be a subset of
\begin{equation}
    \F \coloneqq \{ \rho: \W[\bmz]{\rho} \geq 0 \text{ for all } \bmz \in \cal{P}_d\}.
\end{equation}
Our focus is on states with negativity, so the particular choice of free states is not critical for our analysis. The remaining component that defines any magic theory is the set of free quantum operations. The most basic assumption we require on free operations is that they send any free state to another free state.

@@

We can go further, and show that in fact a range of R\'{e}nyi entropies remain both well-defined and meaningful in the Wigner representation. The $\alpha$--R\'{e}nyi entropy $H_\alpha(\p)$ is defined as~\cite{renyi_1960}
\begin{equation}
	H_\alpha(\p) \coloneqq \frac{1}{1-\alpha} \log \sum_i p_i^\alpha.
\end{equation}
To extend to quasi-distributions we must ensure the following: (a) the entropy is well-defined on quasi-distributions, and (b) obeys monotonicity under the majorization pre-order. In Appendix~\ref{app:single-shot} we prove the following.
\begin{theorem}\label{thm:HSchur} 
	If $\alpha = \frac{2a}{2b-1}$ for positive integers $a,b$ with $a \geq b$, then $H_\alpha(\w)$ is well-defined on the set of quasi-distributions, and if $\w \succ \w'$ for two quasi-distributions $\w, \w'$ then $H_\alpha (\w) \leq H_\alpha(\w')$.
\end{theorem}
This monotonicity provides a meaningful extension of entropies to quasi-distributions. The unusual form of the $\alpha$ parameter is required so that the entropy function is well-defined and real-valued for all $\w$. The allowed values of $\alpha$ are dense in $\alpha >1$, however, while R\'{e}nyi entropies on probabilities can also be used for $0 < \alpha <1$, it turns out that this parameter regime is no longer monotonic in terms of the majorization pre-order for quasi-distributions (see Appendix~\ref{app:single-shot} for details).

@@

We can now apply this to obtain a family of magic distillation bounds based on R\'{e}nyi divergences.
\begin{theorem}\label{thm:entropic_bounds}
	Consider a general magic state distillation protocol on odd prime dimension qudits, that converts a magic state $\rho^{\otimes n} \longrightarrow \E(\rho^{\otimes n})=\rho'^{\otimes m}$and let $\tau$ be any full-rank stabilizer reference state on a qudit. Then the distillation rate $R \coloneqq m/n$ is upper bounded as,
	\begin{equation}
		R \leq \frac{D_{\alpha}(W_\rho \hspace{1pt}||\hspace{1pt} W_\tau)}{\rlap{\raisebox{-0.2ex}{$\widetilde{\phantom{D}}$}}D_\alpha( \rho', \tau')},
	\end{equation}
	where $\alpha = \frac{2a}{2b-1}$ for any positive integers $a,b$ with $a \geq b$ and the average divergence
	\begin{equation}
\widetilde{D}_\alpha( \rho', \tau') \coloneqq \frac{1}{m} D_\alpha (W_{\rho'^{\otimes m}} \hspace{1pt}||\hspace{1pt} W_{\tau'_m}),
\end{equation}
between the output magic state $\rho'^{\otimes m}$ and $\tau'_m = \E(\tau^{\otimes n})$.
\end{theorem}
The proof is given in Appendix~\ref{app:single-shot}.

>>>>

Work towards fault-tolerance is increasingly bridging the gap between abstract theory and experiment. Extensive work is being done on error mitigation~\cite{jones_2012, Li_2017, Temme_2017, Endo_2018, McClean_2017}, and the incorporation of hardware physics into the theoretical models~\cite{Kandala_2019, holmes_resource_2019, Colless_2018, song2018quantum, Bravyi_2021}. For example, the XZZX code~\cite{bonilla_ataides_xzzx_2021} is a variant of the surface code that incorporates noise bias explicitly and has been shown to attain the hashing bound of random codes~\cite{Bennett_1996}. In this work, we develop a framework to analyse magic state distillation protocols where explicit physics of the system is incorporated in the distillation bounds. This is achieved by considering how a given magic state protocol transforms a \emph{pair} of quantum states -- one being a noisy magic state, and the other a stabilizer state that is distinguished by the physics of the system (for example a state at a characteristic temperature, or the maximally mixed state) and acts as a reference state for the protocol. The magic distillation bounds can then be expressed in terms of the physics of the reference state (e.g.~free energy changes in the case of temperature). Such bounds are of potential interest in assessing how physical features such as temperature, noise biases or fixed-point structure associated with restricted gate-sets constrain distillation protocols~\cite{Tuckett_2019, Aliferis_2008, Stephens_2013, Li_2015, Babbush_2018, Guillaud_2019, Fowler_2019}.

@@

The phase-point operators form an orthogonal operator basis with respect to the Hilbert-Schmidt inner product, as discussed in Appendix~\ref{app:wigner}.
Therefore, any quantum state $\rho$ can be expressed as a linear combination of them, $\rho = \sum_{\bmz \in \cal{P}_d} \W[\bmz]{\rho} A_{\bmz}$, where the coefficients $\W{\rho}(\z)$ give the Wigner distribution of state $\rho$,
\begin{equation}\label{eq:wstate}
    \W[\bmz]{\rho} \coloneqq \frac{1}{d}\tr[A_{\bmz} \rho].
\end{equation}
For any quantum state $\rho$, the Wigner distribution $W_\rho(\bmz)$ is readily seen to be a $d^2$-dimensional quasi-distribution over $\cal{P}_d$. More precisely, $W_\rho(\bmz)$ is a real-valued function on $\P_d$ with the normalisation property $\sum_{\bmz} W_\rho(\bmz) = 1$ (see Appendix~\ref{app:wigner} for details). Moreover, the above Wigner representation has recently been shown to be the only non-contextual quasi-probability representation of odd-dimensional stabilizer theories for odd prime dimensions~\cite{schmid2021noncontextual}.

@@

@@

Within the Wigner representation for odd prime dimension, it is well-known that all positively represented states used in Clifford circuits admit an efficient classical simulation~\cite{cit:mari}, so negativity is a necessary resource for universal fault-tolerant quantum computing~\cite{cit:veitch}. Therefore, the free states in any magic theory are required to be a subset of
\begin{equation}
    \F \coloneqq \{ \rho: \W[\bmz]{\rho} \geq 0 \text{ for all } \bmz \in \cal{P}_d\}.
\end{equation}
Our focus is on states with negativity, so the particular choice of free states is not critical for our analysis. The remaining component that defines any magic theory is the set of free quantum operations. The most basic assumption we require on free operations is that they send any free state to another free state.

@@

We can go further, and show that in fact a range of R\'{e}nyi entropies remain both well-defined and meaningful in the Wigner representation. The $\alpha$--R\'{e}nyi entropy $H_\alpha(\p)$ is defined as~\cite{renyi_1960}
\begin{equation}
	H_\alpha(\p) \coloneqq \frac{1}{1-\alpha} \log \sum_i p_i^\alpha.
\end{equation}
To extend to quasi-distributions we must ensure the following: (a) the entropy is mathematically well-defined on quasi-distributions, and (b) obeys monotonicity under the majorization pre-order. In Appendix~\ref{app:single-shot} we prove the following.
\begin{theorem}\label{thm:HSchur} 
	If $\alpha = \frac{2a}{2b-1}$ for positive integers $a,b$ with $a \geq b$, then $H_\alpha(\w)$ is well-defined on the set of quasi-distributions, and if $\w \succ \w'$ for two quasi-distributions $\w, \w'$ then $H_\alpha (\w) \leq H_\alpha(\w')$.
\end{theorem}
This monotonicity provides a meaningful extension of entropies to quasi-distributions. The unusual form of the $\alpha$ parameter is required so that the entropy function is well-defined and real-valued for all $\w$. The allowed values of $\alpha$ are dense in $\alpha >1$, however, while R\'{e}nyi entropies on probabilities can also be used for $0 < \alpha <1$, it turns out that this parameter regime is no longer monotonic in terms of the majorization pre-order for quasi-distributions (see Appendix~\ref{app:single-shot} for details).

@@

We can now apply this to obtain a family of magic distillation bounds based on R\'{e}nyi divergences.
\begin{theorem}\label{thm:entropic_bounds}
	Consider a general magic state distillation protocol on odd prime dimension qudits, that converts a magic state $\rho^{\otimes n} \longrightarrow \E(\rho^{\otimes n})=\rho'^{\otimes m}$and let $\tau$ be any full-rank stabilizer reference state on a qudit. Then the distillation rate $R \coloneqq m/n$ is upper bounded as,
	\begin{equation}
		R \leq \frac{D_{\alpha}(W_\rho \hspace{1pt}||\hspace{1pt} W_\tau)}{\rlap{\raisebox{-0.2ex}{$\widetilde{\phantom{D}}$}}D_\alpha( \rho', \tau')},
	\end{equation}
	where $\alpha = \frac{2a}{2b-1}$ for any positive integers $a,b$ with $a \geq b$ and the average divergence per qudit
	\begin{equation}
\widetilde{D}_\alpha( \rho', \tau') \coloneqq \frac{1}{m} D_\alpha (W_{\rho'^{\otimes m}} \hspace{1pt}||\hspace{1pt} W_{\tau'_m}),
\end{equation}
between the output magic state $\rho'^{\otimes m}$ and $\tau'_m = \E(\tau^{\otimes n})$.
\end{theorem}
The proof is given in Appendix~\ref{app:single-shot}.
